---
title: "Demographic data"
author: "Zuzie"
date: "2023-06-26"
output: word_document
---

Loading the packages and data:

```{r setup, include = FALSE}
library(tidyverse)
slsr <- read.csv("/Users/Zuzanna_Bien/Desktop/ACF Public Health/Cognitive impairment in stroke/SLSR-CI/SLSR_raw_data/ds_202305.csv")

```

---
Demographic data 

1. Age 

Outliers: there are 4 values <17 (including 2 values which are negative). 
There are also 3 values above 110 (including 2 >200 and 1 >900). 

This can be seen on the boxplot below: 

```{r boxplot, echo = FALSE}
slsr  %>% ggplot (aes(x = age)) + geom_boxplot() 
```
To remove biologically implausible ages:
```{r age outliers, echo= FALSE}
#to remove biologically implausible ages 
slsr <- slsr %>% mutate(age = ifelse((age < 0 | age >110), NA,  age))
```
  
Repeating the boxplot now doesn't throw out an error:
```{r boxplot repeat, echo = FALSE}
slsr  %>% ggplot (aes(x = age)) + geom_boxplot() 
```

2. Ethnicity 

`r sum(is.na(slsr$eth6cat))` patients were missing Ethnicity data. 

3. Employment 

`r sum(is.na(slsr$socat))` patients were missing Social Class data.
`r sum(is.na(slsr$employ))` patients were missing Employment data.

Employment status: 
1=full time  
2=part time 
3=unemployed and looking 
4=unable due to disability/ ill health 
5=retired 
6=carer 
8=student

```{r employment}
table(slsr$employ)
```
4. Scan type 
0=None 1=CT 2=MRI 3=CT & MRI 4=CT or MRI

```{r scan type}
table(slsr$scantyp)
```

5. Assistance pre-stroke 

1=No 2=Yes

```{r pre-stroke assistance}
print(sum(is.na(slsr$assist)))
table(slsr$assist)
```

